---
title: A guide to Matching estimators, in R and with DAGs - Part 2
author: Francisco Yirá
date: '2022-03-11'
slug: matching-in-r-part-2
cover: "images/clones_orphan_black.jpg"
useRelativeCover: true
isMath: "true"
categories:
  - causal-inference
  - data-science
  - R
tags:
  - causal-inference-the-mixtape
  - books
  - dags
  - potential-outcomes
  - the-effect
  - matching
  - english-only-ds
---

Welcome to the second part of this article about Matching estimators. If you haven't read the first part, I recommend you to do so, since this second part will build on top of that, but here is a quick summary in case you don't have time for reading another whole article:

The first part covered the conditional independence assumption, which is like a "weak version" of the independence we find in fully randomised trials. It states that the treatment assignment is random inside subgroups of the data with the same values of some covariates X, despite not being random in the data as a whole. For example, you could have an experiment with kids in schools where the treatment is randomised, but different schools have different proportions of treatment assignment. There the school would be our X (a confounder and covariate), and there would be independence between the treatment and (TALK ABOUT potential outcomes).

## Matching vs. Regression: What's the difference?

If you're more familiar with DAGs, you may have realised that, DAG-wise, the contexts in which we can use Matching and Multiple Linear Regression for estimating causal effects are the same. Both require the Conditional Independence Assumption (CIA) to hold, which translates to being able to close all the backdoors by conditioning on a set of observable variables (which we call X).

Put another way, if you need to condition on the covariates X to isolate the causal effect, you could either do Matching based on X or to add X as controls to a regression ($Y = D + X$).

So, how to choose between these two techniques? Let's look into the differences between these two methods to answer that question.

### Regression is parametric, Matching is non-parametric

Regression requires us to assume a functional form that links the the treatment (D) and the covariates (X) with the outcome (Y). We call this kind of model "parametric" because it has parameters that are going to be estimated, which are usually known as "the betas" in the regression context.

On the other hand, Matching provides more freedom by not imposing any functional form on the variables. The only two assumptions it relies on are the CIA (which is required by regression too), and common support (which is necessary to obtain an estimate).

For example, if your regression model is $Y = D + X_1 + X_2$ (where $X_1$ and $X_2$ are covariates and observable confounders), you're imposing a specification that assumes:

-   Independence between the treatment effect and the covariates values.
-   Linear and additive effects of the covariates on the outcome (the "beta" of $X_1$ is the same regardless of the value of $X_2$, and vice-versa)..

These assumptions could be true or not, and if they are not true, then our estimate may be biased due to misspecification.

Meanwhile, if we did Matching using $X_1$ and $X_2$, the estimated counterfactual for each unit would be just the (average) actual outcome of the matched unit(s), so there would be no need to make any functional form assumptions. The relationships between the covariates, the treatment and the outcome could be highly non-linear, and the Matching estimator would still work just fine, as long as common support and the CIA holds[^1].

[^1]: It's worth mentioning that regression could also handle non-linear relationships, but we should have to "declare" such non-linearity in the model specification.

Let's see an example with code:

```{r, message=FALSE, warning=FALSE}
library(tidyverse)

df <- tibble(
  x = runif(1000, -1, 4),
  prob_d = ifelse(x > 0.5 & x < 2.5, 0.1, 0.9),
  d = rbinom(1000, 1, prob_d),
  noise = rnorm(1000, sd = 0.1),
  treat_effect = 1,
  y = sin(x) + d*treat_effect + noise
) %>% 
  mutate(d_factor = factor(d, levels=c(0,1), labels=c("Untreated", "Treated")))

ggplot(df,
       aes(x, y,
           color = d_factor)) +
  geom_point() + 
  labs(color = "Treatment status")
```

```{r}
linear_model1 <- lm(y ~ d + x, data = df)

summary(linear_model1)
```

```{r}
confint(linear_model1)
```

Here `x` is a confounder, because it affects both the outcome and the probability of receiving the treatment, and multiple linear regression can't help us because both relationships are non-linear, so the final estimate is heavily biased (despite including the only confounder as control in the regression). Sad!

Now let's use Matching with the same data. Here we're going to use a Matching estimator implemented by a CRAN R package instead of coding it ourselves[^2]:

[^2]: Which is what we should do most of the time. The R functions of the previous blog post had only pedagogical purposes, and in general is better to leverage implementations of statistical methods available in CRAN packages. Not only are they less likely to have errors, but also they have extra goodies such as standard errors calculation and nice `print` or `summary` methods.

```{r}
library(Matching)
M <- Match(df$y, df$d, df$x, M = 5, Weight = 2, caliper = 0.05)
summary(M)
```

As expected, the Matching estimate is much closer to the true treatment effect (and its confidence interval includes the true value) because it's not relying on false linearity assumptions[^3].

[^3]: The linear regression fan club may object that a dataset like this is very unlikely to be found in the "real world", and that it was literally manufactured with the goal of fooling linear regression (which is true). What's more, linear regression is technically *capable* of handling datasets like this, given that we include the non-linear relationships in the specification (for example, `lm(y ~ d + sin(x), data = df)` returns a treatment effect estimate close to 1). But the point is that we should critically assess if we have good enough reasons to assume a specific functional form between the variables, and realise that if we don't, then a non-parametric approach such as Matching may be a better option.

Here is a illustration that makes it clearer what each of the estimators are doing and why the linear regression estimate is so off[^4]:

[^4]: Yes, KNN regression is not technically the same as matching, but for the effect of this visualisation the KNN algorithm illustrates well what matching is doing.

```{r, message=FALSE, warning=FALSE, fig.height=2.5, out.asp=1}
# Matching 
library(FNN)
knn1 <- knn.reg(
  train = dplyr::select(df, x, d),
  y = df$y,
  test = dplyr::select(df, x, d),
  k = 10,
  algorithm = "brute"
)

df_knn <- df %>% 
  mutate(pred_knn = knn1$pred)

# Regression
df_linear <- df %>% 
  modelr::add_predictions(linear_model1, var = "pred_linear")


# Plots
plot_matching <- 
  ggplot(df_knn) +
  aes(x = x, color = d_factor) +
  geom_point(aes(y = y), alpha = 0.3) +
  geom_line(aes(y = pred_knn), size = 1.5) +
  labs(color = "Treatment status",
       title = "Matching") + theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank()
  )

plot_linear <- 
  ggplot(df_linear) +
  aes(x = x, color = d_factor) +
  geom_point(aes(y = y), alpha = 0.3) +
  geom_line(aes(y = pred_linear), size = 1.5) +
  labs(color = "Treatment status",
       title = "Linear regression")

library(patchwork)

(plot_matching / plot_linear) +
  plot_layout(guides = "collect") + 
  plot_annotation(subtitle = "In each subplot, the estimated ATE is the average distance between the two lines")
    
```

### Common support assumption vs extrapolation

Up to this point, it seems that Matching is winning. Why would we want to assume a specific functional form when we can just... don't?

Sadly, there is no free lunch, and the flexibility of Matching comes at a cost: the common support assumption. In order to get a counterfactual estimate without a parametric model we need to have observations with the same (or similar) covariate values and opposite treatment status in our data. If we don't, Matching will fail.

However, Regression doesn't have such requisite. Instead, it leverages the parametric specification to get estimates by "extrapolation" in the feature space areas where there isn't common support. And if our parametric specification appropriately reflects the data generating process, then regression is the way to go in this case 💃🏽[^5].

[^5]: Big If, though.

Let's see a visual example to better understand this idea.

```{r}
# VO2 max example

```

Here we have a non-linear confounder, just like in the previous example, but without common support, so Matching can't help us. But the regression model is correctly specified and, therefore, it correctly estimates the true ATE.

### Other differences

These differences are mostly consequences of the parametric vs non-parametric dilemma

-   Weighting when we have heterogeneous treatment effects

-   Bias vs variance

(weighting on frequency of covariates vs weighting on treatment variance)

And in some contexts, both methods will return the same estimate

```{r}
lm(y ~ d + region, data = customers) %>% 
  summary()
```

## Propensity Scores

Okey, so now we know the

(What it is) + (credits to Rubin)

(propensity score DAG and theorem / balancing property)

(propensity score as dimensionality reduction, curse of dimensionality being avoided)

(common support again)

### Inverse Probability Weighting

(criticism of matching with propensity score)

(estimator for the ATE, and for the ATT)

(visualisations showing the weighting of similar units)

(Problem of sensitivity to extreme values: - trimming - normalised estimator)

## Don't treat propensity scores as a machine learning prediction problem

## How to estimate the variance of these estimators?

### Matching: don't get creative, use proven formulas and software

### IPW: bootstrap

## Double Robust Estimator

(what it is, and the logic behind it)

(example in code)

## Implementations as R packages

(References and so on)
